---
title: "Project 1 - Basic Recommender System"
author: "J. Hamski, V. Nguy"
date: "June 12, 2016"
output: html_document
runtime: shiny
---

```{r load_libraries, warning = FALSE, error = FALSE, message=FALSE, cache = TRUE}
library(recommenderlab)
library(Matrix)
library(dplyr)
library(tidyr)
library(ggplot2)
require(shiny)
library(webshot)

options(scipen=999)
```

# Description
[Last.fm](http://www.lastfm.com), a web application which allows users to track what music they play on their computer or mobile devices, released a listener dataset from 2006. [Dataset information](http://files.grouplens.org/datasets/hetrec2011/hetrec2011-lastfm-readme.txt)


Using Last.fm data we will build a simple recommender system that will recommend music to the users. We will be building this recommender system using collaborative filtering. The system will use an Item-based collaborative filter to recommend music to the users by recommending more artists based on a single artist.


# Dataset
[Last.FM](http://grouplens.org/datasets/hetrec-2011/) data from grouplens.org. The dataset we are interested in is the 'user_artists.dat' data file. This file contains the artists listened by each user. It also provides a listening count for each [user, artist] pair.  

Reading in needed datasets. Artist information will be joined to user-artist dataset for readability
```{r Read_Dataset, cache=TRUE, warnings = FALSE}
user.artists.pairwise <- read.delim('hetrec2011-lastfm-2k/user_artists.dat', sep="\t")

artists <- read.delim('hetrec2011-lastfm-2k/artists.dat', sep="\t") %>%
  select(id, name)
```

#Data Preparation

Converting from Artist IDs to Artist Names for easier interpretation.
```{r join_artist_name, cache=TRUE}
colnames(artists) <- c("artistID", "name")

#note: one duplicate artist
#artists$name[duplicated(artists$name)==TRUE]
user.artists.pairwise <- inner_join(user.artists.pairwise, artists, by="artistID")

user.artists.pairwise <- select(user.artists.pairwise, -artistID)

user.artists.pairwise$name <- strtrim(user.artists.pairwise$name, 100)

user.artists.pairwise[16060,]$name <- NA

ggplot(user.artists.pairwise, aes(x = weight)) + geom_density() + theme_bw() + scale_x_log10()
```

Using the TidyR library. The pairwise data was converted into a sparse matrix. Due to dataset size, we reduced to it's size by eliminating any artist with less than 1000 listens. 
```{r spread_data, cache=TRUE}
num_listens <- 1000

user.artists.matrix <- user.artists.pairwise %>%
  filter(weight > num_listens) %>% #filter artists with less than num_listens listens
  spread(key = name, value = weight) %>%
  as.matrix()

user.artists <- user.artists.matrix[,-1]
dim(user.artists)
```


# Calculating Similarity using the 'recommenderlab' package  

First, we used the recommenderlab package to compute the similiarity matrix for artists. 
```{r reclab_sim_matrix, cache=TRUE}
user.artists.RRM <- new("realRatingMatrix", data = as(user.artists, "CsparseMatrix"))

#reclab.similarity.user <- similarity(user.artists.RRM, method = "cosine", which = "users")
reclab.similarity.item <- similarity(user.artists.RRM, method = "cosine", which = "items")

reclab.sim.item.matrix <- as.matrix(reclab.similarity.item)
```



```{r sim_matrix, cache=TRUE}
#image(as.matrix(reclab.similarity.user))
image(reclab.sim.item.matrix)
```


Let's take a look at artist similar to the band Beach House. Beach House is a fairly obscure indie rock band that plays a sub-genre called 'chill wave'. As expected, the similar artists specified by the similarity matrix tend to be obscure indie rock bands, some of which also play chill wave (e.g. Best Coast).
```{r reclab_recommendations_1, cache=TRUE}
beachhouse.sim.row <- reclab.sim.item.matrix["Beach House",]

beachhouse.similar.listens <- beachhouse.sim.row[complete.cases(beachhouse.sim.row)]

sort(beachhouse.similar.listens, decreasing = TRUE)[1:20]
```


Looking at a more well known rock band, Nine Inch Nails, we tend to find similar artists in the metal genre.
```{r artist_recommendations_2}
nin.sim.row <- reclab.sim.item.matrix["Nine Inch Nails",]

nin.similar.listens <- nin.sim.row[complete.cases(nin.sim.row)]

sort(nin.similar.listens, decreasing = TRUE)[1:20]
```


By searching similar artist to Britney Spears we find some sensible similar artists (e.g. Kylie Minogue, Ashlee Simpson). However, our second highest similarity is a band called "†‡† " or "Ritualz" which makes obscure experimental electronic music. This will warrant more exploration. 
```{r artist_recommendations_2..2}
nin.sim.row <- reclab.sim.item.matrix["Britney Spears",]

nin.similar.listens <- nin.sim.row[complete.cases(nin.sim.row)]

sort(nin.similar.listens, decreasing = TRUE)[1:20]
```

# Calculating Similiarity using built-in Matrix Algebra Functions

```{r, eval=FALSE}

cosineDist <- function(x){
  as.dist(1 - x%*%t(x)/(sqrt(rowSums(x^2) %*% t(rowSums(x^2))))) 
}

cosineDist(user.artists)

image(as.matrix(user.artists))
```

# Calculate Similarity using our own Cosine Distance Algorithm 
```{r custom_similarity_function, eval = F}
euc_distance <- function(x) {
  return (sqrt(sum(x^2,na.rm=TRUE)))
}

cos_sim_matrix <- function(x) {
  # initialize similarity matrix
  m <- matrix(NA, nrow=ncol(x),ncol=ncol(x),dimnames=list(colnames(x),colnames(x)))
  similarity <- as.data.frame(m)
  
  for(i in 1:ncol(x)) {
    for(j in i:ncol(x)) {
      #Take only the columns in which both users ranked the item
      co_rate_1 <- x[which(x[,i] & x[,j]),i]
      co_rate_2 <- x[which(x[,i] & x[,j]),j]  
      #Calculate the Cosine Similarity
      similarity[i,j]= sum(co_rate_1*co_rate_2) / (euc_distance(x[,i]) * euc_distance(x[,j]))
      #Replicate results for the 'other half' of the matrix
      similarity[j,i]=similarity[i,j]        
    }
  }
  return(similarity)
}
```


```{r, eval=F}
#Test case to validate results. Using toy dataset from meetup. Set to false before final rendering.
movie <- read.csv('UBCF.csv', header = TRUE, stringsAsFactors = FALSE)
rownames(movie) <- movie[,1]
movie <- movie[,-1]

#Results should match Item-Item CF tab
cos_sim_matrix(movie)

#Results should match User-User CF tab
cos_sim_matrix(t(movie))
```

```{r custom_sim_matrix, warning=FALSE, eval=F}
#Results should match Item-Item CF tab
custom.similarity.item <- cos_sim_matrix(user.artists)

#Results should match User-User CF tab
#custom.similarity.user <- cos_sim_matrix(t(user.artists))
```

```{r custom_artist_recommendations_1, eval=F}
recommendation.custom <- custom.similarity.item["Beach House",]
recommendation.custom <- as.matrix(t(recommendation.custom))

#Remove records with no score
recommendation.custom <- recommendation.custom[!(apply(recommendation.custom, 1, function(y) any(y == 0))),]

#Remove 'identity recommendation'
recommendation.custom <- recommendation.custom[recommendation.custom<0.99]

sort(recommendation.custom, decreasing = TRUE)
```

```{r custom_artist_recommendations_2, eval=F}
recommendation.custom <- custom.similarity.item["Nine Inch Nails",]
recommendation.custom <- as.matrix(t(recommendation.custom))

#Remove records with no score
recommendation.custom <- recommendation.custom[!(apply(recommendation.custom, 1, function(y) any(y == 0))),]

#Remove 'identity recommendation'
recommendation.custom <- recommendation.custom[recommendation.custom<0.99]

sort(recommendation.custom, decreasing = TRUE)
```


# Artist Recommender

```{r, echo = FALSE}

similarity.recommender <-  as.data.frame(reclab.sim.item.matrix) %>% add_rownames("artist")
```

```{r}

shinyApp(
  
  ui = fluidPage(
    fluidRow(
    selectInput("artist", "Select Artist:", selected = "Beach House",
                choices = colnames(reclab.sim.item.matrix))),
    fluidRow(
      tableOutput("table"))
  ),
  
  server = function(input, output) {
    
    output$table <- renderTable({

      data <- similarity.recommender %>%
        filter(artist == input$artist) %>%
        sort(decreasing = TRUE) %>%
        top_n(20)
      
    })
  },
  
  options = list(height = 500)
)
```