---
title: "Project 1 - Basic Recommender System"
author: "J. Hamski, V. Nguy"
date: "June 12, 2016"
output: pdf_document
---

```{r, warning = FALSE, error = FALSE, message=FALSE}
library(recommenderlab)
library(Matrix)
library(dplyr)
library(tidyr)

options(scipen=999)
```

# Description
Using Last.fm data we will build a simple recommender system that will recommend music to the users. We will be building this recommender system using collaborative filtering. The system will use both a Item-based collaborative filter and User-based collaborative filtering (seperately) to recommend music to the users.

[Dataset information](http://files.grouplens.org/datasets/hetrec2011/hetrec2011-lastfm-readme.txt)

# Dataset
[Last.FM](http://grouplens.org/datasets/hetrec-2011/) data from grouplens.org. The dataset we are interested in is the 'user_artists.dat' data file. This file contains the artists listened by each user. It also provides a listening count for each [user, artist] pair.

```{r}
user.artists.pairwise <- read.delim('hetrec2011-lastfm-2k/user_artists.dat', sep="\t")

artists <- read.delim('hetrec2011-lastfm-2k/artists.dat', sep="\t") %>%
  select(id, name)

head(user.artists.pairwise)
```

## Converting from Artist IDs to Artist Names for easier interpretation
```{r}
colnames(artists) <- c("artistID", "name")
#note: one duplicate artist
#artists$name[duplicated(artists$name)==TRUE]

user.artists.pairwise <- inner_join(user.artists.pairwise, artists, by="artistID")

user.artists.pairwise <- select(user.artists.pairwise, -artistID)

user.artists.pairwise[16060,]$name <- NA
```

#Data Preparation
Using the TidyR library. The pairwise data was converted into a sparse matrix.

```{r}
user.artists.matrix <- user.artists.pairwise %>%
  spread(key = name, value = weight) %>%
  as.matrix()
```

```{r}
user.artists <- user.artists.matrix[,-1]
head(user.artists)[,1:5]
```


For development, work with a smaller matrix. Set eval = FALSE for final rendering. 
```{r, eval=TRUE}
#note - perhaps this should select randomly instead of the first 1000 rows / 500 columns?
#user.artists <- user.artists[1:200, 1:500]

user.artists.colSums <- colSums(user.artists, na.rm=TRUE)

hist(user.artists.colSums)

colnames(user.artists.colSums) <- c("artist", "ColSum")

```


# Calculating Similarity using the 'recommenderlab' package
```{r}
#note this is takes a few minutes to compute
user.artists.RRM <- new("realRatingMatrix", data = as(user.artists, "CsparseMatrix"))

#reclab.similarity.user <- similarity(user.artists.RRM, method = "cosine", which = "users")
reclab.similarity.item <- similarity(user.artists.RRM, method = "cosine", which = "items")

reclab.sim.item.matrix <- as.matrix(reclab.similarity.item)
```

```{r}
image(as.matrix(reclab.similarity.user))
image(reclab.sim.item.matrix)
```

```{r}
#Punk band Agent Orange - we should expect it shares similarity with other punk bands. 
AgentOrange.sim.row <- reclab.sim.item.matrix["Agent Orange",]

AgentOrange.sim.row[complete.cases(AgentOrange.sim.row)]
```


# Calculating Similiarity using built-in Matrix Algebra Functions

```{r, eval=FALSE}

cosineDist <- function(x){
  as.dist(1 - x%*%t(x)/(sqrt(rowSums(x^2) %*% t(rowSums(x^2))))) 
}

cosineDist(user.artists)

image(as.matrix(user.artists))
```



# Calculate Similarity using our own Cosine Distance Algorithm 

